\section{Inverse function for base64 decoder}

It's piece of cake for KLEE to reconstruct input base64 string given just base64 decoder code without corresponding encoder code.
I've copypasted this piece of code from
\url{http://www.opensource.apple.com/source/QuickTimeStreamingServer/QuickTimeStreamingServer-452/CommonUtilitiesLib/base64.c}.

We add constraints (lines 84, 85) so that output buffer must have byte values from 0 to 15.
We also tell to KLEE that the Base64decode() function must return 16 (i.e., size of output buffer in bytes, line 82).

\lstinputlisting[numbers=left]{KLEE/klee_base64.c}

\begin{lstlisting}
% clang -emit-llvm -c -g klee_base64.c
...

% klee klee_base64.bc
KLEE: output directory is "/home/klee/klee-out-99"
KLEE: WARNING: undefined reference to function: klee_assert
KLEE: ERROR: /home/klee/klee_base64.c:99: invalid klee_assume call (provably false)
KLEE: NOTE: now ignoring this error at this location
KLEE: WARNING ONCE: calling external: klee_assert(0)
KLEE: ERROR: /home/klee/klee_base64.c:104: failed external call: klee_assert
KLEE: NOTE: now ignoring this error at this location
KLEE: ERROR: /home/klee/klee_base64.c:85: memory error: out of bound pointer
KLEE: NOTE: now ignoring this error at this location
KLEE: ERROR: /home/klee/klee_base64.c:81: memory error: out of bound pointer
KLEE: NOTE: now ignoring this error at this location
KLEE: ERROR: /home/klee/klee_base64.c:65: memory error: out of bound pointer
KLEE: NOTE: now ignoring this error at this location

...
\end{lstlisting}

We're interesting in the second error, where \TT{klee\_assert()} has been triggered:

\begin{lstlisting}
% ls klee-last | grep err
test000001.user.err
test000002.external.err
test000003.ptr.err
test000004.ptr.err
test000005.ptr.err

% ktest-tool --write-ints klee-last/test000002.ktest
ktest file : 'klee-last/test000002.ktest'
args       : ['klee_base64.bc']
num objects: 1
object    0: name: b'input'
object    0: size: 32
object    0: data: b'AAECAwQFBgcICQoLDA0OD4\x00\xff\xff\xff\xff\xff\xff\xff\xff\x00'
\end{lstlisting}

This is indeed a real base64 string, terminated with the zero byte, just as it's requested by C/C++ standards.
The final zero byte at 31th byte (starting at zeroth byte) is our deed: so that KLEE would report lesser number of errors. % FIXME spelling

The base64 string is indeed correct:

\begin{lstlisting}
% echo AAECAwQFBgcICQoLDA0OD4 | base64 -d | hexdump -C
base64: invalid input
00000000  00 01 02 03 04 05 06 07  08 09 0a 0b 0c 0d 0e 0f  |................|
00000010
\end{lstlisting}

base64 decoder Linux utility I've just run blaming for ``invalid input''---it means the input string is not properly padded.
Now let's pad it manually, and decoder utility will no complain anymore:

\begin{lstlisting}
% echo AAECAwQFBgcICQoLDA0OD4== | base64 -d | hexdump -C
00000000  00 01 02 03 04 05 06 07  08 09 0a 0b 0c 0d 0e 0f  |................|
00000010
\end{lstlisting}

The reason our generated base64 string is not padded is because base64 decoders are usually discards padding symbols (``='') at the end.
In other words, they are not require them, so is the case of our decoder.
Hence, padding symbols are left unnoticed to KLEE.

So we again made \textit{antipode} or \textit{inverse function} of base64 decoder.

