\subsection{List of SMT-solvers}

\begin{itemize}

\item Yices\footnote{\url{http://yices.csl.sri.com/}}, created by Bruno Dutertre et al.

\item Z3\footnote{\url{https://github.com/Z3Prover/z3}},
developed by Leonardo de Moura, Nikolaj Bjorner, Christoph M. Wintersteiger, Lev Nachmanson.

Many examples here uses Python 3.x API for Z3.
Here is how to install it in Ubuntu:

\begin{lstlisting}
sudo apt-get install python3-pip
sudo pip3 install z3-solver
\end{lstlisting}

Or compile latest on Ubuntu:

\begin{lstlisting}
git clone https://github.com/Z3Prover/z3.git
cd z3
git tag
git checkout z3-4.8.4	# or another version
python scripts/mk_make.py --python
cd build
make
sudo make install
\end{lstlisting}

(Unofficial) bindings:
Haskell\footnote{\url{http://hackage.haskell.org/package/z3}},
Racket\footnote{\url{https://github.com/philnguyen/z3-rkt}}.

\item STP\footnote{\url{https://github.com/stp/stp}}, used in KLEE.

\item CVC3/CVC4\footnote{\url{http://cvc4.stanford.edu/}}.

\item Boolector\footnote{\url{http://fmv.jku.at/boolector/}}, developed by Aina Niemetz, Mathias Preiner and Armin Biere.
Known to be fastest bitvector solver.

\item Alt-Ergo\footnote{\url{https://alt-ergo.ocamlpro.com/}}, used in Frama-C.

\item MathSAT\footnote{\url{http://mathsat.fbk.eu/}}. Developed by Alberto Griggio, Alessandro Cimatti and Roberto Sebastiani.

\item veriT\footnote{\url{http://www.verit-solver.org/}}.
Developed by David Déharbe, Pascal Fontaine, Haniel Barbosa.
Lacks bitvectors.

\item toysolver\footnote{\url{https://github.com/msakai/toysolver}} by Masahiro Sakai, written in Haskell.

\item MK85\footnote{\url{https://github.com/DennisYurichev/MK85}}. Created by Dennis Yurichev, as a toy-level bit-blaster.

\item dReal: ``An SMT Solver for Nonlinear Theories of the Reals''
\footnote{\url{http://dreal.cs.cmu.edu}, \url{https://github.com/dreal}}.

\end{itemize}

Something else:

\begin{itemize}

\item PySMT: unified Python interface to many SMT solvers: \url{https://pysmt.readthedocs.io/en/latest/}

\item JavaSMT -- Unified Java API for SMT solvers: \url{https://github.com/sosy-lab/java-smt}

\item jSMTLIB -- Another Java API for SMT solvers: \url{http://smtlib.github.io/jSMTLIB/}

\item SBV: SMT Based Verification in Haskell: \url{http://leventerkok.github.io/sbv/}

\end{itemize}

