\chapter{(Amateur) cryptography}
\label{crypto}

\section{\emph{Professional} cryptography}

Let's back to the method we previously used (\ref{symbolic_exec}) to construct expressions using running Python function.

We can try to build expression for the output of XXTEA encryption algorithm:

\lstinputlisting[style=custompy]{crypto/xxtea.py}

A key is choosen according to input data, and, obviously, we can't know it during symbolic execution, so we leave expression like \TT{k[...]}.

Now results for just one round, for each of 4 outputs:

\lstinputlisting{crypto/1round.txt}

Somehow, size of expression for each subsequent output is bigger. I hope I haven't been mistaken?
And this is just for 1 round.
For 2 rounds, size of all 4 expression is $\approx 970KB$.
For 3 rounds, this is $\approx 115MB$.
For 4 rounds, I have not enough RAM on my computer.
Expressions \emph{exploding} exponentially.
And there are 19 rounds.
You can weigh it.

Perhaps, you can simplify these expressions: there are a lot of excessive parenthesis,
but I'm highly pessimistic, cryptoalgorithms constructed in such a way to not have any spare operations.

In order to crack it, you can use these expressions as system of equation and try to solve it using SMT-solver.
This is called ``algebraic attack''.

In other words, theoretically, you can build a system of equation like this: $MD5(x)=12341234...$,
but expressions are so huge so it's impossible to solve them.
Yes, cryptographers are fully aware of this and one of the goals of the successful cipher is
to make expressions as big as possible, using resonable time and size of algorithm.

Nevertheless, you can find numerous papers about breaking these cryptosystems with reduced number of rounds:
when expression isn't \emph{exploded} yet, sometimes it's possible.
This cannot be applied in practice, but such an experience has some interesting theoretical uses.

\subsection{Attempts to break ``serious'' crypto}

CryptoMiniSat itself exist to support XOR operation, which is ubiquitous in cryptography.

\begin{itemize}
\item Bitcoin mining with SAT solver: \url{http://jheusser.github.io/2013/02/03/satcoin.html}, \url{https://github.com/msoos/sha256-sat-bitcoin}.

\item \href{http://2015.phdays.ru/program/dev/40400/}{Alexander Semenov, attempts to break A5/1, etc. (Russian presentation)}

\item \href{https://yurichev.com/mirrors/SAT_SMT_crypto/thesis-output.pdf}{Vegard Nossum - SAT-based preimage attacks on SHA-1}

\item \href{https://yurichev.com/mirrors/SAT_SMT_crypto/166.pdf}{Algebraic Attacks on the Crypto-1 Stream Cipher in MiFare Classic and Oyster Cards}

\item \href{https://yurichev.com/mirrors/SAT_SMT_crypto/Attacking-Bivium-Using-SAT-Solvers.pdf}{Attacking Bivium Using SAT Solvers}

\item \href{https://yurichev.com/mirrors/SAT_SMT_crypto/Extending_SAT_2009.pdf}{Extending SAT Solvers to Cryptographic Problems}

\item \href{https://yurichev.com/mirrors/SAT_SMT_crypto/sat-hash.pdf}{Applications of SAT Solvers to Cryptanalysis of Hash Functions}

\item \href{https://yurichev.com/mirrors/SAT_SMT_crypto/slidesC2DES.pdf}{Algebraic-Differential Cryptanalysis of DES}

\item Attepmt to break Speck cipher: \url{https://github.com/TechSecCTF/z3_splash_class}.
\end{itemize}

\section{Amateur cryptography}

This is what you can find in serial numbers, license keys, executable file packers, \ac{CTF}, malware, etc.
Sometimes even ransomware (but rarely nowadays, in 2017).

Amateur cryptography is often can be broken using SMT solver, or even KLEE.

Amateur cryptography is usually based not on theory, but on visual complexity:
if its creator getting results which are seems chaotic enough, often, one stops to improve it further.
This is security based not even on obscurity, but on a chaotic mess.
This is also sometimes called ``The Fallacy of Complex Manipulation''
(see \href{https://tools.ietf.org/html/rfc4086}{RFC4086}).

Devising your own cryptoalgorithm is a very tricky thing to do.
This can be compared to devising your own \ac{PRNG}.
Even famous Donald Knuth in 1959 constructed one, and it was visually very complex,
but, as it turns out in practice, it has very short cycle of length 3178.
[See also: The Art of Computer Programming vol.II page 4, (1997).]

The very first problem is that making an algorithm which can generate very long expressions is tricky thing itself.
Common mistake is to use operations like XOR and rotations/permutations, which can't help much.
Even worse: some people think that XORing a value several times can be better, like: $(x \oplus 1234) \oplus 5678$.
Obviously, these two XOR operations (or more precisely, any number of it) can be reduced to a single one.
Same story about applied operations like addition and subtraction---they all also can be reduced to single one.

Real cryptoalgorithms, like IDEA, can use several operations from different groups, like XOR, addition and multiplication.
Applying them all in specific order will make resulting expression irreducible.

When I prepared this article, I tried to make an example of such amateur hash function:

\lstinputlisting[style=customc]{crypto/1.c}

KLEE can break it with little effort.
Functions of such complexity is common in shareware, which checks license keys, etc.

Here is how we can make its work harder by making rotations dependent of inputs,
and this makes number of possible inputs much, much bigger:

\lstinputlisting[style=customc]{crypto/2.c}

Addition (or \href{https://yurichev.com/blog/modulo/}{modular addition}, as cryptographers say) can make thing even harder:

\lstinputlisting[style=customc]{crypto/3.c}

As en exercise, you can try to make a block cipher which KLEE wouldn't break.
This is quite sobering experience.

Another significant property of the \textit{serious} cryptography is:
``The two inputs differ by only a single bit, but approximately half the bits are different in the digests.''
(Alan A. A. Donovan, Brian W. Kernighan ---  The Go Programming Language).

Another easy way to test your algorithm: encrypt numbers starting at 0 and feed the resulting ciphertext to \emph{diehard test}
\footnote{\url{https://en.wikipedia.org/wiki/Diehard_tests}} (like in Counter/CTR encryption mode).
These tests are designed to check \ac{PRNG}s.
In other words, tests shouldn't find any regularities in a list of random numbers, as well as in a ciphertext.

Summary: if you deal with amateur cryptography, you can always give KLEE and SMT solver a try.
Even more: sometimes you have only decryption function, and if algorithm is simple enough,
KLEE or SMT solver can reverse things back.

One amusing thing to mention: if you try to implement amateur cryptoalgorithm in Verilog/VHDL language to run it on \ac{FPGA},
maybe in brute-force way,
you can find that \ac{EDA} tools can optimize many things during synthesis
(this is the word they use for ``compilation'') and can leave cryptoalgorithm much smaller/simpler than it was.
Even if you try to implement \ac{DES} algorithm \emph{in bare metal} with a fixed key,
Altera Quartus can optimize first round of it and make it smaller than others.

\subsection{Bugs}

Another prominent feature of amateur cryptography is bugs.
Bugs here often left uncaught because output of encrypting function visually looked ``good enough'' or ``obfuscated enough'',
so a developer stopped to work on it.

This is especially feature of hash functions, because when you work on block cipher, you have to do two functions
(encryption/decryption), while hashing function is single.

Weirdest ever amateur encryption algorithm I once saw, encrypted only odd bytes of input block, while even bytes
left untouched, so the input plain text has been partially seen in the resulting encrypted block.
It was encryption routine used in license key validation.
Hard to believe someone did this on purpose.
Most likely, it was just an unnoticed bug.

\subsection{XOR ciphers}

Simplest possible amateur cryptography is just application of XOR operation using some kind of table.
Maybe even simpler. This is a real algorithm I once saw:

\begin{lstlisting}
for (i=0; i<size; i++)
    buf[i]=buf[i]^(31*(i+1));
\end{lstlisting}

This is not even encryption, rather concealing or hiding.

Some other examples of simple XOR-cipher cryptoanalysis are present in the ``Reverse Engineering for Beginners''
\footnote{\url{http://beginners.re}} book.

\subsection{Other features}

\textbf{Tables} There are often table(s) with pseudorandom data, which is/are used chaotically.

\textbf{Checksumming} End-users can have proclivity of changing license codes, serial numbers, etc., with a hope
this could affect behaviour of software.
So there is often some kind of checksum: starting at simple summing and \ac{CRC}.
This is close to \ac{MAC} in real cryptography.

\textbf{Entropy level} Maybe (much) lower, despite the fact that data looks random.

\subsection{Examples}

\begin{itemize}

\item A popular FLEXlm license manager was based on a simple amateur cryptoalgorithm
(before they switched to \ac{ECC}), which can be cracked easily.

\item Pegasus Mail Password Decoder: \url{http://phrack.org/issues/52/3.html} -
a very typical example.

\item You can find a lot of blog posts about breaking \ac{CTF}-level crypto using Z3, etc.
Here is one of them: \url{http://doar-e.github.io/blog/2015/08/18/keygenning-with-klee/}.

\item Many examples of amateur cryptography I've taken from an old Fravia website:
\url{https://yurichev.com/mirrors/amateur_crypto_examples_from_Fravia/}.

\item Dmitry Sklyarov's book "The art of protecting and breaking information" (2004, in Russian) has many examples of amateur cryptography and 
misused cryptographical algorithms as well.

\item Reversing the petya ransomware with constraint solvers (Z3)
\footnote{\url{https://0xec.blogspot.com/2016/04/reversing-petya-ransomware-with.html}}.

\end{itemize}

% subsection
\input{crypto/hash/main_EN}
\input{crypto/XOR/main_EN}

\section{Swapping encryption and decryption procedures}

By the way, this is important property of symmetric cryptography: encryption and decryption procedures can be swapped
without loss of security:

\begin{lstlisting}[style=custompy]
# copypasted from https://pypi.org/project/pycrypto/ and reworked:

from Crypto.Cipher import AES

obj = AES.new('This is a key123', AES.MODE_CBC, 'This is an IV456')
message = "Hello, world!!!!" # message's size must be multiple of 16

# decrypt "Hello, world" message:
rubbish = obj.decrypt(message)

print rubbish # rubbish printed

# encrypt it back:

obj2 = AES.new('This is a key123', AES.MODE_CBC, 'This is an IV456')
plaintext_again=obj2.encrypt(rubbish)

print plaintext_again # "Hello, world" printed
\end{lstlisting}

The author saw this in shareware checking license keys, etc, when a key is encrypted (yes) to get information from it.
Perhaps, their key generator decrypted data when putting it to key file.
Hopefully, they swapped functions just by mistake, but it worked, so they left it all as is.

